---
title: "Canada 2018 Data Analysis"
output: html_notebook
---

```{r}
# Load the packages
library("jsonlite", lib.loc="~/anaconda3/lib/R/library")
library("stringr", lib.loc="~/anaconda3/lib/R/library")
library("dplyr", lib.loc="~/anaconda3/lib/R/library")
library("tibble", lib.loc="~/anaconda3/lib/R/library")
library("ggplot2", lib.loc="~/anaconda3/lib/R/library")
library("reshape2", lib.loc="~/anaconda3/lib/R/library")
```



## Import the data and clean it
```{r lapsImportClean}
# Import the data in JSON format
url = "http://ergast.com/api/f1/2018/6/laps.json?limit=2000"
data <- fromJSON(url)

# Get the laps from the data
laps <- data[["MRData"]][["RaceTable"]][["Races"]][["Laps"]]

# Get the times and positions from laps
times_positions <- laps[[1]][["Timings"]]

# Get the lap numbers
lap_numbers <- laps[[1]][["number"]]
lap_numbers <- as.numeric(lap_numbers)

# Convert into a single data frame
times_positions <- do.call("rbind", times_positions)

# Write a function which calculates the time column as an number
time_numeric_seconds <- function(x){
  
  # Get the first character and convert to a time in seconds
  minute <- as.numeric(str_sub(x, 1, 1))
  
  # Get the seconds and milliseconds
  seconds <- as.numeric(str_sub(x, 3))
  
  # Find the time
  time <- 60*minute + seconds
}

# Convert the time column into seconds in numeric form
time_numeric <- sapply(times_positions$time, time_numeric_seconds)

# Bind to the dataframe and drop the original column
times_positions <- times_positions[,-3]
times_positions <- cbind(times_positions, time = time_numeric)

# Convert the position column to numeric
times_positions$position <- as.numeric(times_positions$position)
```

```{r raceResultsImportClean}
# Import the data in JSON format
url = "http://ergast.com/api/f1/2018/6/results.json?limit=2000"
data <- fromJSON(url)

# Get the results dataframe
results <- data[["MRData"]][["RaceTable"]][["Races"]][["Results"]][[1]]
#str(results)

# Extract the driverId column and store as a vector
driverId <- results[["Driver"]][["driverId"]]

# Extract constructor info
constructorId <- results[["Constructor"]][["constructorId"]]

# Create the dataframe of results that I want
df_results <- subset(results, select=c("position", "grid", "laps", "status"))
df_results <- cbind(df_results, driverId)
df_results <- cbind(df_results, constructorId)
df_results$driverId <- as.character(df_results$driverId)
df_results$constructorId <- as.character(df_results$constructorId)
df_results$grid <- as.numeric(df_results$grid)
df_results$position <- as.numeric(df_results$position)
df_results$laps <- as.numeric(df_results$laps)

# Convert data.frame to tibble
results <- as_tibble(df_results)
results
```


## Visualisation
### Prepare data and packages
```{r, echo = FALSE}
# Convert data.frame to tibble
times_positions <- as_tibble(times_positions)

# Add the lap numbers using dplyr
times_positions_laps <- times_positions %>%
 group_by(driverId) %>%
 mutate(lap = row_number())

# Find some lap statistics for all the drivers using dplyr
lap_statistics <- times_positions_laps %>%
  group_by(driverId) %>%
  summarise(medianLapTime = median(time), slowestLapTime = max(time), fastestLapTime = min(time))
```
### Lap Times
#### Bar Plots
```{r}
# Create a function to plot a bar chart with driverId on the y and variable on x
plot_lap_statistics <- function(lap_statistics_df, colName){
  ggplot(lap_statistics_df, aes(x = reorder(driverId, -lap_statistics_df[[colName]]), y = lap_statistics_df[[colName]])) + 
    geom_col() + coord_flip() + labs(x = "driver", y = "time", title = colName)
}

# Plot the max, median and min lap times for each driver
plot_lap_statistics(lap_statistics, "slowestLapTime")
plot_lap_statistics(lap_statistics, "medianLapTime")
plot_lap_statistics(lap_statistics, "fastestLapTime")
```

#### Point Plots
```{r}
# Try and plot all on one plot

# Create a function to plot a bar chart with driverId on the y and variable on x
plot_lap_statistics <- function(lap_statistics_df, colName){
  ggplot(lap_statistics_df, aes(x = reorder(driverId, -lap_statistics_df[[colName]]), y = lap_statistics_df[[colName]])) + 
    geom_point() + coord_flip() + labs(x = "driver", y = "time", title = colName)
}

# Plot the max, median and min lap times for each driver
plot_lap_statistics(lap_statistics, "slowestLapTime")
plot_lap_statistics(lap_statistics, "medianLapTime")
plot_lap_statistics(lap_statistics, "fastestLapTime")
```

#### Point Plot
```{r, echo = FALSE}
# Conver the data from wide format to long format using melt from reshape2
lap_statistics2 <- melt(lap_statistics, id.vars = "driverId", measure.vars = c("fastestLapTime", "medianLapTime", "slowestLapTime"))

# Rename the column names
names(lap_statistics2) <- c("driverId", "lapMeasure", "time")

lap_statistics2
```

```{r}
# Get a sorted list of median laptimes for driverId

# Remove the Slowest Lap Times
lap_statistics3 <- lap_statistics2 %>%
  filter(lapMeasure != "slowestLapTime") %>%
  arrange(time)

# Extract the driverId column for median lap times
lap_statistics4 <- lap_statistics3 %>%
  filter(lapMeasure == "medianLapTime") %>%
  arrange(time)

# Get a list of driverIds in reverse order
driverIdMedian <- rev(lap_statistics4$driverId)
```


```{r}
# Plot the data all on one point plot
ggplot(lap_statistics3, aes(x = factor(lap_statistics3$driverId, levels = driverIdMedian), y = time, color = lapMeasure)) + 
    geom_point() + coord_flip() + labs(x = "driver", y = "time", title = "Fastest and Median Lap Times", subtitle = "Monaco Grand Prix 2018") + 
    scale_fill_discrete(name="Lap Measure",
                         breaks=c("fastestLapTime", "medianLapTime"),
                         labels=c("Fastest Lap Time", "Slowest Lap Time"))
```

#### Drivers
Lets create a function that sorts the data by a vector of driverIds. After the dataframe is sorted then a density plot can be made with one plot for each driver.

```{r}
# Function to filter by drivers vector
filterByDrivers <- function(.driverIds) {
  df <- times_positions_laps %>%
    filter(driverId %in% .driverIds)
  return(df)
}

# Function to plot a histogram for the driver called
plotDensityDriverTimes <- function(.driverIds) {
  
  # Filter by driver
  df <- filterByDrivers(.driverIds)
  
  # Plot a histogram of lap times
  ggplot(df, aes(x = time, colour = driverId)) + geom_density() + coord_cartesian(xlim=c(73, 110))
}
```


##### Ricciardo, Vettel and Hamilton

Lets see how the top 3 finishes compared. It looks like in general that Vettel had the most pace of the top 3 (mode of the density plot is furthest to the left). However, he was stuck behind Ricciardo for most of the race as shown in the similarity between their density plots. Lewis Hamilton was very consistent throughout the race (large mode), however, since he didn't have the outright pace, he couldn't challenge for the lead. 

```{r}

plotDensityDriverTimes(c("ricciardo", "vettel", "hamilton"))

```

### Versus Qualifying Results
How did the driver's compare against their starting positions? The drivers listed on the bottom right corner outperformed relative to their starting positions. This means they gained positions through the race. 

```{r}
# Scatter plot with finish position vs starting position
ggplot(results, aes(x = grid, y = position, label = driverId)) + geom_abline(aes(intercept = 0, slope = 1)) + geom_label(aes(fill = factor(constructorId)), colour = "white", fontface = "bold") 
```


```{r}
results2 <- results %>%
  mutate(startFinishDifference = grid - position)

# Create a bar plot of start finish difference
ggplot(results2, aes(x = reorder(driverId, results2[["startFinishDifference"]]), y = startFinishDifference)) + 
    geom_col(aes(fill = constructorId)) + coord_flip() + labs(x = "driver", y = "Grid Position - Finish Position", title = "Difference Between Start and Finish Position")
```

